---
title: "Vectors and indexing"
draft: true
---

R has a special data structure called a vector. 
A vector is a 1D set of the same type of object.
Most often, a vector will simply be a sequence of numbers.
We can create a sequence of numbers using the `:` operator.

```{r}
numbers <- 1:10
numbers
```

Note that vectors are treated the same way as a single element.
Anything that works on a single number works the same way on a vector.
This is called vectorization.

```{r}
numbers + 5
2 ^ numbers
sin(numbers)
```

We can also create a vector with the `c()` function 
(`c` stands for concatenate, in case you are wondering).

```{r}
concat <- c(4, 17, -1, 55, 2)
concat
```

## Indexing

Often, we don't want to get the entire vector. 
Perhaps we only want a single element, or a set of specific elements.
We do this by indexing (uses the `[]` brackets).

To get the first element of a vector, we could do the following.
In R, array indexes start at 1 - the 1st element is at index 1.
This is different than 0-based languages like C, Python, or Java 
where the first element is at index 0.

```{r}
concat[1]
```

To get other elements, we could do the following:

```{r}
concat[2]  # second element
concat[length(concat)]  # last element
```

Notice that for the second example, we put a function inside the square brackets.
In this case, `length()` is used to get the length of a vector, 
and since the lenght of the vector will be equal to the index of its last element,
this is a nifty way of getting the last element of a vector.

We can actually put anything inside the square brackets. 
Putting another vector inside the brackets gives us multiple values, for instance.

```{r}
concat[1:4]
concat[c(3, 5)]
```

We can use this technique to reassign certain values inside the vector.
For instance, we could change the 3rd and 5th values to 76 with the following code:

```{r}
concat[c(3, 5)] <- 76
concat
```

It's even possible to index outside the bounds of a vector.
Notice that R "fills in the blanks" with `NA` values.
`NA` is R's placeholder for "no data" (since 0 often shows up in real data).

```{r}
concat[10] <- 4.3
concat
```

**DON'T EVER DO THIS.**
Though this is actually valid code in R 
(indexing outside of a vector's size is an error in most other languages),
it comes at a performance cost.
We will go more in-depth as to why later on.


